Supermartingales and Termination Analysis of Statistical PCF

### 2020-05-27 probabilisticProgrammingMartingales.tex
Added new section on antitone-convex ranking function; title and table of contents

\input{Style/comment}

\usepackage{graphicx}

@@

\newcommand{\Real}{\mathbb{R}}
\newcommand\expect[1]{\mathbb{E}[#1]}
\newcommand\set[1]{\{#1\}}
\newcommand\dif{\mathrm{d}}
\newcommand\calF{\mathcal{F}}
\newcommand\calI{\mathcal{I}}
\newcommand\Leb{\mathrm{Leb}}
\newcommand\ndraw[2]{\#\mathrm{draw}_{#1}(#2)}

@@

\usepackage{amsfonts}
\usepackage{amsthm}
\usepackage{amssymb}
\usepackage{verbatim}
\newcommand{\tY}{\textsf{Y}}
\newcommand{\tif}[3]{\textsf{if }#1\textsf{ then }#2\textsf{ else }#3}
\newcommand{\tsample}{\textsf{sample}}
\newcommand{\tscore}{\textsf{score}}
\DeclareMathOperator{\red}{red}
\DeclareMathOperator{\next}{next}

@@

\theoremstyle{remark}
\newtheorem*{remark}{Remark}
\begin{document}

\akr{Following your suggestion, I will be providing a criterion for termination of programs in PPCF \citep{DBLP:journals/pacmpl/EhrhardPT18} based on ranking supermartingales. 
As it's more convenient for this proof, a sampling-based semantics \lo{You need to provide reference(s) for sampling-based semantics.} will be used instead of the original distributional semantics. 
I assume some roughly applicable equivalence is proven somewhere, but it doesn't seem that hard anyway.}
\lo{Why not provide a proof as an appendix?}

@@

\section{Language definition}
The language SPCF is a simply-typed lambda calculus with sampling of real numbers from $[0,1]$ and unbounded scoring, following \cite{MakOP20b}. Types and terms are defined as follows, where $r$ is a real, $x$ is a variable, $f : \mathbb{R}^n \to \mathbb{R}$ is any measurable function and $\Gamma$ is an environment:
\begin{align*}
  & \text{types } A, B ::= \textsf{R}  \mid  A \to B \\
  & \text{values } v ::= \lambda x.s  \mid  \underline{r} \\
  & \text{terms } s, t ::= v  \mid  x  \mid  t_1 t_2  \mid  \underline{f}(s_1,\dots ,s_n)  \mid  \tY s  \mid  \tif{s < 0}{t_1}{t_2}  \mid  \tsample  \mid  \tscore(s)
\end{align*}
\begin{align*}
  \frac{}{\Gamma ; x:A \vdash x:A} \qquad
  \frac{\Gamma ; x:A \vdash s : B}{\Gamma \vdash \lambda x.s : A \to B} \qquad
  \frac{}{\underline{r} : \textsf{R}} \qquad
  \frac{\Gamma \vdash s:A \to B \quad \Gamma \vdash t : A}{\Gamma \vdash s \, t : B} \\ \\
  \frac{\Gamma \vdash s_1:\textsf{R} \dots \Gamma \vdash s_n:\textsf{R}}{\Gamma \vdash \underline{f}(s_1,\dots,s_n) : \textsf{R}} \ f : \mathbb{R}^n \to \mathbb{R} \qquad
  \frac{\Gamma \vdash s : (A \to B) \to (A \to B)}{\Gamma \vdash \tY s : (A \to B)} \\ \\
  \frac{\Gamma \vdash c : \textsf{R} \quad \Gamma \vdash s_1 : A \quad \Gamma \vdash s_2 : A}{\Gamma \vdash \tif{c < 0}{s_1}{s_2} : A} \qquad
  \frac{}{\Gamma \vdash \tsample : \textsf{R}} \qquad
  \frac{\Gamma \vdash s : \textsf{R}}{\Gamma \vdash \tscore (s) : \textsf{R}}
\end{align*}

@@

\subsection{Alternative Semantics}
\akr{This version of the semantics is much more convenient for allowing multiple different reduction orders. I don't currently have a proof of the equivalence of this with the more usual semantics. I expect that that would be somewhat complicated, but not terribly difficult. I don't know whether a semantics like this has already been defined elsewhere. I just thought I'd write it out so you can see it and in case I use it later. If I do end up actually using it I'll write it up nicer, but hopefully this is at least enough for you to understand what I mean.

@@

\section{Constructing Ranking Functions}
Although rankability implies almost-sure termination, the converse does not hold in general. For example,
\begin{equation}
\tif{{-}\tsample < 0}{\underline{0}}{(\tY \lambda x. x) \; \underline 0}
\end{equation}
terminates in 3 steps with probability 1, but isn't rankable because $(\tY \lambda x. x) \underline 0$ is reachable, although that has probability 0. 
Not only is this counterexample a.s.t, it's positively almost surely terminating i.e.~the expected time to termination is finite.

@@

Even in this simple case, defining a ranking function explicitly is awkward because of the number of cases, although in most cases, because the value need only be greater than or equal to that of the next term in sequence, it suffices to take the ranking function as taking only 3 distinct values.
\lo{This is only true for partial ranking function.}

@@

Define $(\next(N,s),\_) = \red^n(N,s)$ for the least $n \geq 0$ such that it's in the domain of $f_1$, and $g(N,s) = \left | \{m < n \mid \red^m(N,s) \text{ is of the form } (E[\tY N'],s') \} \right |$. 
The function $\next$ is well-defined (i.e.~$n$ is finite) for all $N \in Rch(M)$ by induction on the path from $M$ to $N$, by the third condition on partial ranking functions. Define $f_2(N) = \int_S f_1(\next(N,s)) + g(N,s) \, \mu(\mathrm d s)$. The (total) function $f_2$ agrees with $f$ on $f$'s domain, and it is a ranking function on $M$ (in fact, the least ranking function of which $f$ is a restriction).
\end{proof}

@@

Maybe I have misread / misunderstood the preceding, but I don't see why $h$ is a restriction of a ranking function. 
Suppose $h$ is a restriction of $H$.
Consider the reduction sequence (writing $\Xi = (\tY f, n. n \oplus_{0.5} f(n+1))$):
\[
\Xi \, 0 
\xrightarrow{\tY} 
0 \oplus_{0.5} \Xi \, 1 
\to 
\Xi \, 1
\xrightarrow{\tY} 
1 \oplus_{0.5} \Xi \, 2
\to
\Xi \, 2
\xrightarrow{\tY} 
2 \oplus_{0.5} \Xi \, 3
\]
Since $H(\Xi \, 0) = 2$, we must have 
$H(\Xi \, 1) \leq 1$ and $H(\Xi \, 2) \leq 0$, contradicting the restriction assumption.
}

>>>>

\input{Style/comment}
\usepackage{calculation}
\usepackage{graphicx}

@@

\newcommand{\Real}{\mathbb{R}}
\newcommand{\nnReal}{\mathbb{R}_{{\geq}0}}
\newcommand{\pReal}{\mathbb{R}_{{>}0}}

\newcommand\expect[1]{\mathbb{E}[#1]}
\newcommand\set[1]{\{#1\}}
\newcommand\dif{\mathrm{d}}
\newcommand\calF{\mathcal{F}}
\newcommand\calI{\mathcal{I}}
\newcommand\Leb{\mathrm{Leb}}
\newcommand\ndraw[2]{\#\mathrm{draw}_{#1}(#2)}

@@

\usepackage{amsfonts}
\usepackage{amsthm}
\usepackage{amssymb}
\usepackage{verbatim}
\newcommand{\tY}{\textsf{Y}}
\newcommand{\tif}[3]{\textsf{if }#1\textsf{ then }#2\textsf{ else }#3}
\newcommand{\tsample}{\textsf{sample}}
\newcommand{\tscore}{\textsf{score}}
\DeclareMathOperator{\red}{red}
\DeclareMathOperator{\nnext}{next}

@@

\theoremstyle{remark}
\newtheorem*{remark}{Remark}

\title{Supermartingales and Termination Analysis of Statistical PCF}

\begin{document}

\maketitle

\lo{Tentative title}

\begin{abstract}
\lo{TODO}
\end{abstract}

\iffalse
\akr{Following your suggestion, I will be providing a criterion for termination of programs in PPCF \citep{DBLP:journals/pacmpl/EhrhardPT18} based on ranking supermartingales. 
As it's more convenient for this proof, a sampling-based semantics \lo{You need to provide reference(s) for sampling-based semantics.} will be used instead of the original distributional semantics. 
I assume some roughly applicable equivalence is proven somewhere, but it doesn't seem that hard anyway.}
\lo{Why not provide a proof as an appendix?}
\fi

\tableofcontents

@@

\section{Statistical PCF}

\subsection{Syntax of SPCF}

The language SPCF is a simply-typed lambda calculus with sampling of real numbers from $[0,1]$ and unbounded scoring, following \cite{MakOP20b}. Types and terms are defined as follows, where $r$ is a real, $x$ is a variable, $f : \mathbb{R}^n \to \mathbb{R}$ is any measurable function and $\Gamma$ is an environment:
\begin{align*}
  & \text{types } A, B ::= \textsf{R}  \mid  A \to B \\
  & \text{values } v ::= \lambda x.s  \mid  \underline{r} \\
  & \text{terms } s, t ::= v  \mid  x  \mid  t_1 t_2  \mid  \underline{f}(s_1,\dots ,s_n)  \mid  \tY s  \mid  \tif{s < 0}{t_1}{t_2}  \mid  \tsample  \mid  \tscore(s)
\end{align*}
\begin{align*}
  \frac{}{\Gamma ; x:A \vdash x:A} \qquad
  \frac{\Gamma ; x:A \vdash s : B}{\Gamma \vdash \lambda x.s : A \to B} \qquad
  \frac{}{\underline{r} : \textsf{R}} \qquad
  \frac{\Gamma \vdash s:A \to B \quad \Gamma \vdash t : A}{\Gamma \vdash s \, t : B} \\ \\
  \frac{\Gamma \vdash s_1:\textsf{R} \dots \Gamma \vdash s_n:\textsf{R}}{\Gamma \vdash \underline{f}(s_1,\dots,s_n) : \textsf{R}} \ f : \mathbb{R}^n \to \mathbb{R} \qquad
  \frac{\Gamma \vdash s : (A \to B) \to (A \to B)}{\Gamma \vdash \tY s : (A \to B)} \\ \\
  \frac{\Gamma \vdash c : \textsf{R} \quad \Gamma \vdash s_1 : A \quad \Gamma \vdash s_2 : A}{\Gamma \vdash \tif{c < 0}{s_1}{s_2} : A} \qquad
  \frac{}{\Gamma \vdash \tsample : \textsf{R}} \qquad
  \frac{\Gamma \vdash s : \textsf{R}}{\Gamma \vdash \tscore (s) : \textsf{R}}
\end{align*}

@@

\section{Alternative Semantics}
\akr{This version of the semantics is much more convenient for allowing multiple different reduction orders. I don't currently have a proof of the equivalence of this with the more usual semantics. I expect that that would be somewhat complicated, but not terribly difficult. I don't know whether a semantics like this has already been defined elsewhere. I just thought I'd write it out so you can see it and in case I use it later. If I do end up actually using it I'll write it up nicer, but hopefully this is at least enough for you to understand what I mean.

@@

\section{Constructing Ranking Functions}
Although rankability implies almost-sure termination, the converse does not hold in general. For example,
\begin{equation}
\tif{{-}\tsample < 0}{\underline{0}}{(\tY \lambda x. x) \; \underline 0}
\label{ex:0 probability reachable}
\end{equation}
terminates in 3 steps with probability 1, but isn't rankable because $(\tY \lambda x. x) \underline 0$ is reachable, although that has probability 0. 
Not only is this counterexample AST, it's positively almost surely terminating i.e.~the expected time to termination is finite.

@@

Even in this simple case, defining a ranking function explicitly is awkward because of the number of cases, although in most cases, because the value need only be greater than or equal to that of the next term in sequence, it suffices to take the ranking function as taking only 3 distinct values.
(We will explain why later.)

@@

Define $(\nnext(N,s),\_) = \red^n(N,s)$ for the least $n \geq 0$ such that it's in the domain of $f_1$, and $g(N,s) = \left | \{m < n \mid \red^m(N,s) \text{ is of the form } (E[\tY N'],s') \} \right |$. 
The function $\nnext$ is well-defined (i.e.~$n$ is finite) for all $N \in Rch(M)$ by induction on the path from $M$ to $N$, by the third condition on partial ranking functions. Define $f_2(N) = \int_S f_1(\nnext(N,s)) + g(N,s) \, \mu(\mathrm d s)$. The (total) function $f_2$ agrees with $f$ on $f$'s domain, and it is a ranking function on $M$ (in fact, the least ranking function of which $f$ is a restriction).
\end{proof}

@@

Maybe I have misread / misunderstood the preceding, but I don't see why $h$ is a restriction of a ranking function. 
Suppose $h$ is a restriction of $H$.
Consider the reduction sequence (writing $\Xi = (\tY f, n. n \oplus_{0.5} f(n+1))$):
\[
\Xi \, 0 
\xrightarrow{\tY} 
0 \oplus_{0.5} \Xi \, 1 
\to 
\Xi \, 1
\xrightarrow{\tY} 
1 \oplus_{0.5} \Xi \, 2
\to
\Xi \, 2
\xrightarrow{\tY} 
2 \oplus_{0.5} \Xi \, 3
\]
Since $H(\Xi \, 0) = 2$, we must have 
$H(\Xi \, 1) \leq 1$ and $H(\Xi \, 2) \leq 0$, contradicting the restriction assumption.
}
\akr{
In this case, $H(0 \oplus_{0.5} \Xi 1) \leq 1$, but that term reduces to either $0$ or $\Xi 1$ therefore the actual condition is that $0.5 H(0) + 0.5 H(\Xi 1) \leq 1$, which is possible.
}
\lo{
  OK. I see it now.
}

\section{Antitone-convex ranking functions}

\begin{example}
\label{ex:ac-ranking}
\begin{enumerate}
\item 1D unbiased random walk
\[
M_1 = 
\big(\tY \, f \, n \, . \, 
\tif{n = 0}{0}{f \, (n - 1) \oplus_{1/2} f \, (n + 1)}\big) \, 1
\]

\item 1D biased (away from 0) random walk
\[
M_2 = 
\big(\tY \, f \, n \, . \, 
\tif{n = 0}{0}{f \, (n - 1) \oplus_{1/3} f \, (n + 1)}\big) \, 10
\]

\item 1D biased (towards 0) random walk
\[
M_3 = 
\big(\tY \, f \, n \, . \, 
\tif{n = 0}{0}{f \, (n - 1) \oplus_{2/3} f \, (n + 2)}\big) \, 10
\]

%While {x > 0} do {x := x-1 \oplus_2/3 x:= x+1}

%M2 = While {x > 0} do {x := x-1 \oplus_2/3 x:= x+2}
\end{enumerate}
\end{example}

\paragraph{Problem.} Find a generalised notion of ranking function 
%--- call it \emph{antitone-convex rankable} --- 
so that $M_1$ becomes rankable, and then prove it sound i.e.~rankable in this generalised sense implies AST.

\paragraph{Idea.}

The definition of antitone-convex ranking function $f$ (say) for $M \in \Lambda^0$ is the same as that of ranking function except that in the case of $\tY$-redex, 
we require the existence of an antitone (meaning: $r < r'$ implies $\epsilon(r) \geq \epsilon(r')$) and convex function $\epsilon : \pReal \to \pReal$ such that the ranking function $f :\mathit{Rch}(M) \to \nnReal$ satisfies
\[
f(E[R’]) \leq f(E[R]) - \epsilon(f(E[R])) 
\]
where $R \to R’$ is the $\tY$-redex rule.

\lo{An aside: This CBV $\tY$-rule seems cleaner:
\[
\big(\tY f^{A \to B} \, x^A \, . \, \theta^B \big) \, v \to \theta[(\tY f \, x \, . \, \theta) / f, v / x].
\]
We assume $\tY f^{A \to B} \, x^A \, . \, \theta^B$ is a value.}

\begin{definition}
We say that a supermartingale $(Y_n)_{n \in \omega}$ adapted to filtration $(\calF_n)_{n \in \omega}$ is \emph{antitone-convex strict} if $Y_n \geq 0$, and there exists an antitone and convex measurable function $\epsilon : \nnReal \to \pReal$ satisfying
\[
\expect{Y_{n+1} \mid \calF_n} \leq Y_n - (\epsilon \circ Y_n) \cdot 1_{\set{Y_n > 0}}
\]
%We say that the supermartingale is \emph{antitone-convex strict} if $\epsilon$ is further assumed to be convex.
\end{definition}

\begin{theorem}
\label{thm:a-c strict}
Let $(Y_n)_{n \in \omega}$ be a antitone-convex strict supermartingale via function $\epsilon : \nnReal \to \pReal$, and define stopping time $T(s) := \min \set{n \mid Y_n(s) = 0}$. 
Then 
\begin{enumerate}
\item $T < \infty$ almost surely, and 
\item $\expect{T} \leq \dfrac{\expect{Y_0}}{\epsilon_0}$ where $\epsilon_0 := \epsilon(\expect{Y_0})$.
\end{enumerate}
\end{theorem}

\begin{proof}
\emph{Claim}: for all $n \in \omega$,
\(
\expect{Y_n} \leq \expect{Y_0} - \epsilon_0 \cdot \Big(\sum_{i=0}^{n-1} \mathbb{P}[Y_i > 0]\Big).
\)
We prove the claim by induction on $n$.
The base case is trivial.
For the inductive case, we have
\iffalse
\everymath{\displaystyle}
\begin{calculation}
\int_A\nu(x,\dif y)\;\frac{\dif \nu'}{\dif\mu}(x,y)  
  \step{decomposition of $\nu = \kact{\mu }{ f}$} 
\int_A\mu(x,\dif y)\;f(x,y)\cdot \frac{\dif \nu'}{\dif\mu}(x,y)
  \step{$f=1$ outside $\infty[\nu]$} 
\int_A\mu(x,\dif y)\;\frac{\dif \nu'}{\dif\mu}(x,y)
  \step{RN Theorem for $\sigma$-finite measures}
\int_A\nu'(x,\dif y)
  \step{definition of integral}
\nu'(x,A).
\end{calculation}
\fi
{%\everymath{displaystyle}
\begin{calculation}
\expect{Y_{n+1}}
  \step[=]{definition of conditional expectation}
\expect{\expect{Y_{n+1} \mid \calF_n}}
  \step[\leq]{linearity of expectation}
\expect{Y_n} - \int \epsilon \circ Y_n \cdot 1_{\set{Y_n > 0}} \, \dif \mu
\; = \;
\expect{Y_n} - \int_{\set{Y_n > 0}} \epsilon \circ Y_n \, \dif \mu
  \step[\leq]{convexity of $\epsilon$, and Jensen's inequality}
\expect{Y_n} - \epsilon \big(\int_{\set{Y_n > 0}} Y_n \, \dif \mu\big)
  \step[\leq]{by assumption $Y_n \geq 0$}
\expect{Y_n} - \epsilon (\expect{Y_n})
  \step[\leq]{$\mathbb{P}[Y_n > 0] \leq 1$}
\expect{Y_n} - \epsilon (\expect{Y_n}) \cdot \mathbb{P}[Y_n > 0]
  \step[\leq]{$\epsilon$ is antitone, and $\epsilon_0 = \expect{Y_0} \geq \expect{Y_n}$}
\expect{Y_n} - \epsilon_0 \cdot \mathbb{P}[Y_n > 0]
  \step[\leq]{induction hypothesis}
%\expect{Y_0} - \epsilon_0 \cdot \Big(\sum_{i=0}^{n-1} \mathbb{P}[Y_i > 0]\Big) - \epsilon_0 \cdot \mathbb{P}[Y_n > 0]
%\step[\leq]{}
\expect{Y_0} - \epsilon_0 \cdot \Big(\sum_{i=0}^{n} \mathbb{P}[Y_i > 0]\Big)
\end{calculation}
}

\iffalse
\begin{align*}
\expect{Y_{n+1}}
&=
\expect{\expect{Y_{n+1} \mid \calF_n}}\\
&\leq
\expect{Y_n} - \expect{\epsilon \circ Y_n} \cdot \mathbb{P}[Y_n > 0]\\
&\leq
\expect{Y_0} - \epsilon_0 \cdot \Big(\sum_{i=0}^{n-1} \mathbb{P}[Y_i > 0]\Big)
- \expect{\epsilon \circ Y_n} \cdot \mathbb{P}[Y_n > 0]\\
&\leq
\expect{Y_0} - \epsilon_0 \cdot \Big(\sum_{i=0}^{n} \mathbb{P}[Y_i > 0]\Big)
\end{align*}
\fi

It follows from the claim that $\sum_{i=0}^{\infty} \mathbb{P}[Y_i > 0]$ converges; hence $\lim_{n \to \infty} \mathbb{P}[Y_n > 0] = 0$ and so $\mathbb{P}[T < \infty] = 1$.

For 2, it suffices to prove: if $T < \infty$ a.s., then $\expect{T} = \sum^\infty_{i=1} \mathbb{P}[T \geq i]$.
\end{proof}

\paragraph{Results to prove:}
\begin{enumerate}
\item (AST Soundness) If a closed SPCF term $M$ is antitone-convex rankable, then $T^{\tY}_M < \infty$ a.s.~(equivalently $M$ is AST), and $\expect{T^{\tY}_M} < \infty$.

\item A partial ranking function theorem for antitone-convex ranking functions.
\end{enumerate}

\paragraph{\Cref{ex:ac-ranking} revisited}

Programs $M_1$ and $M_3$ are AST.
To prove the existence of antitone-convex ranking functions, we appeal \lo{(hopefully)} to a partial ranking function theorem for antitone-convex ranking functions.

\lo{NOTE. Suppose a function $f : E \to \Real$ from a convex subset $E \subseteq \Real^n$ has continuous second-order partial derivatives. Then it is convex iff its Hessian is positive semidefinite.}

For $M_1$, consider the function $g_1(n) := \frac{2^n - 1}{2^{n-1}}$.
Note that $(g_1 (n))_{n \geq 0}$ satisfies the recurrence relation
\[
z_0 = 0, \qquad
z_n > 1/2 \cdot z_{n-1} - 1/2 \cdot z_{n+1} \quad (n \geq 1)
\]
and set the antitone and convex function $\epsilon_1$ to be $x \mapsto g_1(x) - 1/2 \cdot g_1(x-1) - 1/2 \cdot g_1(x+1) = \frac{1}{2^{x+1}}$.

For $M_3$, we use the function $g_3(n) := \frac{3^n - 1}{3^{n-1}}$, and set $\epsilon_3$ to be $x \mapsto g_3(x) - 2/3 \cdot g_3(x-1) - 1/3 \cdot g_3(x+2) = 
\frac{28}{3^{x+2}}$.

\subparagraph{Questions.} 
\begin{enumerate}
\item Can we remove the convexity assumption?
\item Complete method for AST?

\lo{Example~\ref{ex:0 probability reachable} seems somewhat artificial: its validity depends on $I$ being the closed interval. 
There is no reason why we can't define $I$ to be the open interval $(0, 1)$, in which case, $Rch(M)$ would consist of all terms that are reachable from $M$ with non-zero probability.}
\end{enumerate}
